---
title: "Peregrine Falcons"
author: "Andy Teucher"
date: "November 21, 2014"
output: md_document
---

```{r}
source("header.R")
```

```{r}
data(peregrine)

model1 <- jags_model("model {
  alpha ~ dnorm(0, 100^-2)
  beta ~ dnorm(0, 100^-2)
  sigma ~ dunif(0, 100)
  
  for(i in 1:length(Pairs)) {
    ePairs[i] <- alpha + beta * Year[i]
    Pairs[i] ~ dnorm(ePairs[i], sigma^-2)
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    prediction[i] <- alpha + beta * Year[i]
  }
}",
select_data = c("Pairs", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

coef(analysis1)
plot(analysis1)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
                     
```

## Exercise 16 (rerun with year not centered):

```{r}
select_data(model1) <- c("Pairs", "Year")

analysis2 <- jags_analysis(model1, data = peregrine)
coef(analysis2)

pred2 <- predict(analysis2)

gp <- gp %+% pred2
gp
```

## Exercise 17 (use a log-link function so can't include 0 Pairs):

```{r}
data(peregrine)

model1 <- jags_model("model {
  alpha ~ dnorm(0, 100^-2)
  beta ~ dnorm(0, 100^-2)
  sigma ~ dunif(0, 100)
  
  for(i in 1:length(Pairs)) {
    log(ePairs[i]) <- alpha + beta * Year[i] # Add log
    Pairs[i] ~ dnorm(ePairs[i], sigma^-2)
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    log(prediction[i]) <- alpha + beta * Year[i] # Add log
  }
}",
select_data = c("Pairs", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

coef(analysis1)
plot(analysis1)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
```

## Exercise 18 (use poisson distribution):

```{r}
data(peregrine)

model1 <- jags_model("model {
  alpha ~ dnorm(0, 100^-2)
  beta ~ dnorm(0, 100^-2) 
  # Removed sigma because not used in the model (in pois mean = var)
  
  for(i in 1:length(Pairs)) {
    log(ePairs[i]) <- alpha + beta * Year[i] # Add log
    Pairs[i] ~ dpois(ePairs[i]) # use a Poisson distribution because counts
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    log(prediction[i]) <- alpha + beta * Year[i] # Add log
  }
}",
select_data = c("Pairs", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

coef(analysis1)
plot(analysis1)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
```

### A bit of off-roading to look at over-dispersed poisson:

This adds a new parameter with a gamma distribution to introduce more variation
in the point estimates

```{r}
data(peregrine)

model1 <- jags_model("model {
  alpha ~ dnorm(0, 100^-2)
  beta ~ dnorm(0, 100^-2) 
  sDispersion ~ dunif(0,5)
  
  for(i in 1:length(Pairs)) {
    eDispersion[i] ~ dgamma(1 / sDispersion^2, 1 / sDispersion^2)
    log(ePairs[i]) <- alpha + beta * Year[i] # Add log
    Pairs[i] ~ dpois(ePairs[i] * eDispersion[i]) # include dispersion
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    log(prediction[i]) <- alpha + beta * Year[i] # Add log
  }
}",
select_data = c("Pairs", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

coef(analysis1)
plot(analysis1)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
```

## Exercise 19:

Add a second-order (quadratic) polynomial to the overdispersed model

```{r}
data(peregrine)

model1 <- jags_model("model {
  alpha ~ dnorm(0, 100^-2)
  beta ~ dnorm(0, 100^-2)
  beta2 ~ dnorm(0, 100^-2)
  beta3 ~ dnorm(0, 100^-2)
  sDispersion ~ dunif(0,5)
  
  for(i in 1:length(Pairs)) {
    eDispersion[i] ~ dgamma(1 / sDispersion^2, 1 / sDispersion^2)
    log(ePairs[i]) <- alpha + beta * Year[i] + beta2 * Year[i]^2 + beta3 * Year[i]^3
    Pairs[i] ~ dpois(ePairs[i] * eDispersion[i]) # include dispersion
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    log(prediction[i]) <- alpha + beta * Year[i] + beta2 * Year[i]^2 + beta3 * Year[i]^3
  }
}",
select_data = c("Pairs", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

coef(analysis1)
plot(analysis1)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
```

Note that the support for the `sDispersion` parameter is decreasing as the third order polynomial is explaining more of the variation. The percent error is still relatively low (~50, wich is < 80), so we will keep it.

## Exercise 21:

```{r}
pred2 <- predict(analysis1, newdata = data.frame(Year = as.integer(2006)))
```

The third-order polynomial predicts that there will be `r pred2$estimate` in 2006.

## State-Space Population Growth Models

```{r}
data(peregrine)

model6 <- jags_model("model {
  mean_r ~ dnorm(0, 1^-2)
  sd_r ~ dunif(0, 1)
  logN[1] ~ dnorm(0, 10^-2)
  
  for(i in 2:nYear) {
    r[i-1] ~ dnorm(mean_r, sd_r^-2)
    logN[i] <- logN[i-1] + r[i-1]
  }
  
  for(i in 1:length(Pairs)) {
    Pairs[i] ~ dpois(exp(logN[Year[i]]))
  }
  
  logN1 <- logN[1]
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    log(prediction[i]) <- logN[Year[i]]
  }
}",
select_data = c("Pairs", "Year"),
random_effects = list(r = "Year", logN = "Year"))

## Need Year as a factor for the loop
peregrine$Year <- factor(peregrine$Year)
analysis_ss <- jags_analysis(model6, data = peregrine)
coef(analysis_ss)

prediction <- predict(analysis_ss)

gp <- ggplot(data = prediction, 
             aes(x = as.integer(as.character(Year)), y = estimate))
gp <- gp + geom_point(data = dataset(analysis_ss), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
```

## Exercise 22:

```{r}
## Predict as percent change since 1970
prediction <- predict(analysis_ss, base = data.frame(Year = as.factor(1970)))

gp <- ggplot(data = prediction, 
             aes(x = as.integer(as.character(Year)), y = estimate))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "% Change since 1970")
gp <- gp + expand_limits(y = 0)

print(gp)
```

The estimated percent change in 2003 since 1970 is `r prediction$estimate[prediction$Year == 2003]` (95% CI: `r prediction$lower[prediction$Year == 2003]` to `r prediction$upper[prediction$Year == 2003]`)

## Exercise 23:

Predict years up to 2008:

```{r}
data(peregrine)

## Add the years you want to predict for
peregrine_padded <- left_join(data.frame(Year = min(peregrine$Year):2008), 
                       peregrine)

## Need Year as a factor for the loop
peregrine_padded$Year <- factor(peregrine_padded$Year)
analysis_pred <- jags_analysis(model6, data = peregrine_padded)
coef(analysis_pred)

prediction_pred <- predict(analysis_pred)

gp <- ggplot(data = prediction_pred, 
             aes(x = as.integer(as.character(Year)), y = estimate))
gp <- gp + geom_point(data = dataset(analysis_pred), aes(y = Pairs))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = 0)

print(gp)
```

## Exercise 25: Breeding Success

Add a second-order polynial to original linear model

```{r}
data(peregrine)

peregrine$Proportion <- peregrine$R.Pairs / peregrine$Pairs

model1 <- jags_model("model {
  alpha ~ dnorm(0, 1^-2)
  beta ~ dnorm(0, 1^-2)
  beta2 ~ dnorm(0, 1^-2) ## Added beta2
  sigma ~ dunif(0, 1)
  
  for(i in 1:length(Proportion)) {
    eProportion[i] <- alpha + beta * Year[i] + beta2*Year[i]^2 ## Add the quadratic term
    Proportion[i] ~ dnorm(eProportion[i], sigma^-2)
  }
}",
derived_code = "data {
  for(i in 1:length(Proportion)) {
    prediction[i] <- alpha + beta * Year[i] + beta2*Year[i]^2 ## Quadratic here too
  }
}",
select_data = c("Proportion", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Proportion))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = c(0, 1))

print(gp)
```

But there is a problem: the polynomial could allow expected values of < 0 or > 1.
Use a logistic link function...

## Exercise 26:

```{r}
data(peregrine)

peregrine$Proportion <- peregrine$R.Pairs / peregrine$Pairs

model1 <- jags_model("model {
  alpha ~ dnorm(0, 1^-2)
  beta ~ dnorm(0, 1^-2)
  beta2 ~ dnorm(0, 1^-2) ## Added beta2
  sigma ~ dunif(0, 1)
  
  for(i in 1:length(Proportion)) {
    logit(eProportion[i]) <- alpha + beta * Year[i] + beta2*Year[i]^2
    Proportion[i] ~ dnorm(eProportion[i], sigma^-2)
  }
}",
derived_code = "data {
  for(i in 1:length(Proportion)) {
    logit(prediction[i]) <- alpha + beta * Year[i] + beta2*Year[i]^2 ## Quadratic here too
  }
}",
select_data = c("Proportion", "Year+"))

analysis1 <- jags_analysis(model1, data = peregrine)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Proportion))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = c(0, 1))

print(gp)

## Look at tails to see it now doesn't go to zero:
pred <- predict(analysis1, data.frame(Year = 1940:2025))
gp %+% pred

```

## Exercise 27:

```{r}
data(peregrine)

peregrine$Proportion <- peregrine$R.Pairs / peregrine$Pairs

model1 <- jags_model("model {
  alpha ~ dnorm(0, 1^-2)
  beta ~ dnorm(0, 1^-2)
  beta2 ~ dnorm(0, 1^-2) ## Added beta2
  
  for(i in 1:length(Pairs)) {
    logit(eProportion[i]) <- alpha + beta * Year[i] + beta2*Year[i]^2
    R.Pairs[i] ~ dbin(eProportion[i], Pairs[i])
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    logit(prediction[i]) <- alpha + beta * Year[i] + beta2*Year[i]^2
  }
}",
select_data = c("Year+", "Pairs", "R.Pairs"))

analysis1 <- jags_analysis(model1, data = peregrine)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Proportion))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = c(0, 1))

print(gp)
```

## Exercise 29: Add overdispersion to binomial distribution

```{r}
data(peregrine)

peregrine$Proportion <- peregrine$R.Pairs / peregrine$Pairs

model1 <- jags_model("model {
  alpha ~ dnorm(0, 1^-2)
  beta ~ dnorm(0, 1^-2)
  beta2 ~ dnorm(0, 1^-2)
  sDispersion ~ dunif(0,5) ## Add prior for overdispersion parameter
  
  for(i in 1:length(Pairs)) {
    eDispersion[i] ~ dnorm(0, sDispersion^-2)
    logit(eProportion[i]) <- alpha + beta * Year[i] + beta2*Year[i]^2 + eDispersion[i]
    R.Pairs[i] ~ dbin(eProportion[i], Pairs[i])
  }
}",
derived_code = "data {
  for(i in 1:length(Pairs)) {
    logit(prediction[i]) <- alpha + beta * Year[i] + beta2*Year[i]^2
  }
}",
select_data = c("Year+", "Pairs", "R.Pairs"))

analysis1 <- jags_analysis(model1, data = peregrine)
coef(analysis1)

prediction <- predict(analysis1)

gp <- ggplot(data = prediction, aes(x = Year, y = estimate))
gp <- gp + geom_point(data = dataset(analysis1), aes(y = Proportion))
gp <- gp + geom_line()
gp <- gp + geom_line(aes(y = lower), linetype = "dashed")
gp <- gp + geom_line(aes(y = upper), linetype = "dashed")
gp <- gp + scale_x_continuous(name = "Year")
gp <- gp + scale_y_continuous(name = "Pairs")
gp <- gp + expand_limits(y = c(0, 1))

print(gp)
```

The percent error in the coefficient table for sDispersion is relatively low, so unlikely to include zero.  This means it's probably important.

## Add overdispersion correction to AR model:

```{r}
model2 <- jags_model("model {
  theta[1] ~ dnorm(0, 2^-2)
  sigma ~ dunif(0, 2)
  for(i in 2:length(R.Pairs)) {
    theta[i] ~ dnorm(theta[i-1], sigma^-2)
  }
  sDispersion ~ dunif(0, 5)
  
  for(i in 1:length(R.Pairs)) {
    eDispersion[i] ~ dnorm(0, sDispersion^-2)
    logit(eProportion[i]) <- theta[i] + eDispersion[i]
    R.Pairs[i] ~ dbin(eProportion[i], Pairs[i])
  }
}",
derived_code = "data {
  for(i in 1:length(R.Pairs)) {
    logit(prediction[i]) <- theta[Year[i]]
  }
}",
random_effect = list(theta = "Year"),
select_data = c("R.Pairs", "Pairs", "Year"))

data(peregrine)
peregrine$Year <- factor(peregrine$Year)

analysis2 <- jags_analysis(model2, data = peregrine)

prediction <- predict(analysis2)

gp <- gp + aes(as.integer(as.character(Year)))
gp <- gp %+% prediction

print(gp)
```

